Understanding Traffic Patterns and Median Speeds: Insights from Toronto’s Speed Display Program

### 2024-12-12 paper/paper.qmd
added figures, fixed citation, added data cleaning section

---
title: "Understanding Traffic Patterns and Median Speeds: Insights from Toronto’s Speed Display Program"
subtitle: "Northbound and Southbound Traffic Show Higher Speeds; High-Speed Vehicles Raise Median Speeds"
author: Xingjie Yao
thanks: "Code and data are available at: [https://github.com/Stella41603/Toronto-Speeding-Analysis.git](https://github.com/Stella41603/Toronto-Speeding-Analysis.git)."
date: today
date-format: long
abstract: "This study investigates how traffic direction and high-speed vehicles affect median speeds at locations with speed display signs in Toronto. Northbound and Southbound traffic showed higher median speeds than Eastbound, and more high-speed vehicles raised the median speed. These findings highlight how road design and driver behavior influence traffic patterns. The results inform strategies to improve road safety by addressing high-speed driving and targeting high-risk areas."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

For this analysis, we used the `R` programming language [@citeR] to explore data from the Mobile Watch Your Speed Program (WYSP) – Speed Summary in Toronto. The dataset, sourced from Toronto’s Open Data Portal [@opendatatoronto; @toronto_mobile_wysp], provides a summary of vehicle speeds observed at temporary speed display sign installations across the city. These signs are part of an initiative to remind motorists to obey speed limits by displaying their real-time speeds. Key aspects analyzed include the median speed observed during each installation, the count of vehicles exceeding 100 km/h, and the cardinal direction of traffic flow.

A variety of `R` packages were employed to streamline data manipulation, modeling, and visualization. The `tidyverse` suite served as the backbone for data processing, enabling efficient and reproducible workflows [@tidyverse]. File management was handled with `here`, ensuring consistent and reliable dataset access [@citehere]. Data cleaning was facilitated by `janitor`, which prepared the dataset for analysis [@citejanitor], while `testthat` validated the simulated and processed datasets, ensuring workflow integrity [@testthat]. Bayesian modeling, conducted using `rstanarm`, provided robust inference and clear probabilistic interpretations [@rstanarm]. Model summaries were generated with `modelsummary` for concise, intuitive reporting [@modelsummary]. Additionally, `arrow` ensured fast and memory-efficient handling of the dataset, critical for its scale and complexity [@citearrow]. Best practices in workflow organization, as outlined in @tellingstories, ensured transparency and reproducibility, integrating data cleaning, modeling, and visualization into a unified pipeline. This structured approach allowed for meaningful insights while supporting validation and future research extensions.

@@

@@

```{r}
#| label: fig-bound
#| echo: false
#| warning: false
#| message: false
#| fig-cap: The distribution of the median speeds, measured in kilometers per hour (km/h), for vehicles traveling in each cardinal direction Eastbound (EB), Northbound (NB), Southbound (SB), and Westbound (WB), at monitored locations during the installation of traffic signs.

ggplot(analysis_data, aes(x = direction, y = pct_50, fill = direction)) +
  geom_boxplot() +
  labs(
    x = "Direction",
    y = "Median speed (km/h)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@@

```{r}
#| label: fig-speeding
#| fig-cap: Illustrating the relationship between the count of vehicles observed exceeding 100 km/h and the median speed, measured in kilometers per hour, for each cardinal direction, Eastbound (EB), Northbound (NB), Southbound (SB), and Westbound (WB).
#| echo: false
#| warning: false
#| message: false

@@

# Additional data details

>>>>

---
title: "Understanding Traffic Patterns and Median Speeds: Insights from Toronto’s Speed Display Program"
subtitle: "Northbound and Southbound Traffic Show Higher Speeds; High-Speed Vehicles Raise Median Speeds"
author: Xingjie Yao
thanks: "Code and data are available at: [https://github.com/Stella41603/Toronto-Speeding-Analysis.git](https://github.com/Stella41603/Toronto-Speeding-Analysis.git)."
date: today
date-format: long
abstract: "This study investigates how traffic direction and high-speed vehicles affect median speeds at locations with speed display signs in Toronto. Northbound and Southbound traffic showed higher median speeds than Eastbound, and more high-speed vehicles raised the median speed. These findings highlight how road design and driver behavior influence traffic patterns. The results inform strategies to improve road safety by addressing high-speed driving and targeting high-risk areas."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

For this analysis, we used the `R` programming language to explore data from the Mobile Watch Your Speed Program (WYSP) – Speed Summary in Toronto [@citeR]. The dataset, sourced from Toronto’s Open Data Portal, provides a summary of vehicle speeds observed at temporary speed display sign installations across the city [@opendatatoronto; @toronto_mobile_wysp]. These signs are part of an initiative to remind motorists to obey speed limits by displaying their real-time speeds. Key aspects analyzed include the median speed observed during each installation, the count of vehicles exceeding 100 km/h, and the cardinal direction of traffic flow.

A variety of `R` packages were employed to streamline data manipulation, modeling, and visualization. The `tidyverse` suite served as the backbone for data processing, enabling efficient and reproducible workflows [@tidyverse]. File management was handled with `here`, ensuring consistent and reliable dataset access [@citehere]. Data cleaning was facilitated by `janitor`, which prepared the dataset for analysis, while `testthat` validated the simulated and processed datasets, ensuring workflow integrity [@citejanitor; @testthat]. Bayesian modeling, conducted using `rstanarm`, provided robust inference and clear probabilistic interpretations [@rstanarm]. Model summaries were generated with `modelsummary` for concise, intuitive reporting [@modelsummary]. Additionally, `arrow` ensured fast and memory-efficient handling of the dataset, critical for its scale and complexity [@citearrow]. Best practices in workflow organizatio ensured transparency and reproducibility, integrating data cleaning, modeling, and visualization into a unified pipeline [@@tellingstories]. This structured approach allowed for meaningful insights while supporting validation and future research extensions.

## Data source

@@

@fig-medianspeed displays the distribution of median speeds (50th percentile speeds) for vehicles across various monitored locations and directions (Eastbound, Northbound, Southbound, and Westbound). The x-axis represents the median speeds, measured in kilometers per hour (km/h), while the y-axis represents the count of observations falling within each speed range. The distribution is roughly bell-shaped, peaking between 30 and 50 km/h, indicating that most median speeds fall within this range. The data also shows a tapering off towards both extremes, with very few median speeds below 10 km/h or above 60 km/h. This pattern suggests that typical vehicle flow during the monitored period was moderate, with a few outliers exhibiting unusually low or high speeds. The histogram highlights the overall consistency in traffic behavior, but the presence of extreme values may indicate occasional disruptions or anomalies in traffic flow.

```{r}
#| label: fig-medianspeed
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Distribution of median speeds (km/h) for vehicles traveling in monitored directions during the installation of traffic signs. The distribution shows a concentration of median speeds between 30 and 50 km/h, with few instances of very low or high median speeds.

ggplot(analysis_data, aes(x = pct_50)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(
    x = "Median Speed (km/h)",
    y = "Count"
  ) +
  theme_minimal()
```

@@

```{r}
#| label: fig-bound
#| echo: false
#| warning: false
#| message: false
#| fig-cap: The distribution of median speeds (km/h) for vehicles traveling in each cardinal direction (EB, NB, SB, WB) at monitored locations during the installation of traffic signs. The results reveal variability in median speeds, with notable clustering around 50 km/h and some outliers indicating unusually high or low speeds.

ggplot(analysis_data, aes(x = direction, y = pct_50, fill = direction)) +
  geom_violin(trim = FALSE, alpha = 0.6) +  
  geom_jitter(width = 0.1, alpha = 0.5, color = "black", size = 1.5) + 
  labs(
    x = "Direction",
    y = "Median speed (km/h)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@@

```{r}
#| label: fig-speeding
#| fig-cap: The relationship between the count of vehicles observed exceeding 100 km/h and the median speed, measured in kilometers per hour, for each cardinal direction: Eastbound (EB), Northbound (NB), Southbound (SB), and Westbound (WB). Higher counts of vehicles exceeding 100 km/h are generally associated with higher median speeds.
#| echo: false
#| warning: false
#| message: false

@@

# Data cleaning procedure

The dataset was processed to focus on essential variables related to traffic patterns. The relevant columns for analysis—traffic direction, the count of vehicles traveling over 100 km/h, and the median speed—were selected. To ensure data consistency, only records corresponding to the four primary travel directions (Eastbound, Northbound, Southbound, and Westbound) were retained. Additionally, records where the count of high-speed vehicles exceeded 500 were excluded to mitigate the impact of extreme values. Any rows containing missing values in these selected columns were removed to ensure completeness. The cleaned dataset was then standardized by ensuring consistent column naming conventions and saved in a format suitable for further analysis.
